Leadership across cultures and contexts: an exploratory analysis

### 2020-05-12 leadership_across_cultures_contexts.Rmd
Minor edits in discussion.

<!-- DISCUSSES PROSOCIALITY FEATURE -->
**Prosociality**. The prosociality feature of leader qualities, including generosity, interpersonal and oratory skills, and charisma as well as the role of cultural values shaping leadership styles, including proper behavior and having a favorable personality, generally correspond to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and more recently emphasized in evolutionary psychology models of leadership [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @howell_role_2005]. The importance of prosociality  also corresponds to the @henrich_big_2015 Big-man mechanism, which suggests positive assortment of followers to prosocial leaders can facilitate the evolution of cooperative social norms and prestige-based leadership.
<!-- A few variables did not seem to fit prosocial theme, however, such as *Physical health*, *Physical formidable*, and *Drug consumption* (although drug use is often an important social activity). -->

<!-- DISCUSSES COMPETENCIES FEATURE -->
**Leader capacities**. The individual-level competencies feature was comprised of variables associated with dominance-based leadership, including coercive authority and aggressiveness [@tiger_imperial_1971; @barkow_prestige_1975; @barkow_darwin_1989; @henrich_evolution_2001; @cheng_two_2013] as well as a group that included qualities of knowledgeable and intelligent and experienced and accomplished, traits that have been emphasized by the prestige theory [@henrich_evolution_2001; @cheng_two_2013] and the more general computational services theory [@hagen_leadership_2019; @garfield_evolutionary_2019], and which are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport)). Interestingly, being feared and having supernatural qualities clustered more closely with the prestige-style traits than with the dominance-style traits. There is likely a strong connection between shamanism and ritual-based influence with both prestige and dominance styles of leadership. Ritual and supernatural capacities may represent a dimension of leader qualities in which dominance and prestige are coupled forming an alternative style of leadership, as shamans often provide a necessary and valued public good, have special expertise and abilities, are feared and achieve or maintain influence via threats or use of direct and/or indirect aggression. Artistic performance is also implicated in this style of leadership and ceremonial, artistic performances by shamans are common displays which often incorporate the superhuman abilities of shamans [@singh_cultural_2017].

@@

<!-- DISCUSSES SUCCESS FEATURE -->
**Leader success**. Leaders, probably as a consequence of their abilities to provide social benefits (e.g., resolving conflicts and being generous) and/or impose social costs (e.g., punishing and aggressiveness), are also characterized by their material, social, and reproductive success (bottom cluster Figure \@ref(fig:cluster)A), which is consistent with much of the theoretical literature focused on associated outcomes of leadership and social hierarchy [e.g., @von_rueden_mens_2016; @bowles_emergence_2010; @glowacki_role_2013]. Several of these measures of success often co-occur. Among the Trobriands, horticulturalists in Papua New Guinea, @ol06-037 [p. 47] describes relationships between polygynous marriages, increased wealth, and widespread social contacts:

@@

<!-- DISCUSSES BOTH MANAGEMENT FEATURES -->
**Leaders are managers**. In the absence of state-institutions local leaders are often tasked with managing various economic systems and providing social services. Within the domain of leader management, there are likely to be two primary features of human leader functions: strategizing and organizing. Our results correspond very closely to what @van_vugt_cognitive_2007 define as *strategic leadership*, where leaders increase the pay-offs of another individuals behavior thereby incentivizing followership versus *coordinating leadership*, where individuals bound by group challenges must effectively organize collective effort.

@@

# Concluding remarks: Beyond the non-WEIRD 
<!-- NOT SURE OF TITLE -- WANT TO EMPHASIZE NON WEIRD DIVERSITY AND ALSO MENTION BACK TO WERID FINDINGS BRIEFLY -->
In recognizing the limitations of most extant psychological, sociological, and managerial data, many social scientists are now diversifying their samples across nations, cultures, and contexts to better understand *human* psychology, behavior, and culture. We caution that the WEIRD people problem in behavioral science cannot be remedied simply by a non-WEIRD solution. As our results, as well as much of anthropology and ethnography, demonstrate, within the populations that are now often coined a non-WEIRD, there is enormous diversity. Furthermore, larger order categories such as "subsistence strategy" are probably less relevant for understanding behavioral diversity than more salient group contexts. Despite limitations, the ethnographic record present an invaluable opportunity to explore patterns of human diversity and evolutionary theorists and researchers looking beyond WEIRD samples should take special interest in such methods.

We found leaders in the ethnographic record are portrayed as generally prosocial, serve to resolve conflicts and organize cooperation, and receive material, social, and reproductive benefits. These results conform to the characteristics of leaders that are emphasized in the literature on the evolution of leadership and provide broad cross-cultural support for several influential theories, such as the prestige-dominance model, the collective action model, the service-for-prestige model, and the computational services model [for review, see @garfield_evolutionary_2019].

While there is a evidence of a "dominant" style of leadership involving coercive authority, confidence and aggression, as well as a "prestige" style involving special knowledge, experience, and innovation, we suggest rethinking models of leadership styles and focusing on inter-individual variation in specific competencies. There is likely to be substantial overlap in physical, spiritual, cognitive, and other prosocial endowments and the degree to which a leader relies on any will be shaped by their individual levels of such capacities, the presence and capacities of their rivals, as well as group context and cultural pressures. For example, widespread preferences for egalitarianism and social equality are likely strong drivers of leader preferences across diverse societies [@cheng_dominance_2019], but shamanism and ritual based contexts represent a social domain in which ideologies support a style of leadership fusing both dominance and prestige-based competencies.

>>>>

<!-- DISCUSSES PROSOCIALITY FEATURE -->
**Prosocial norms**. The prosociality feature of leader qualities, including generosity, interpersonal and oratory skills, and charisma as well as the role of cultural norms in shaping leadership styles, including proper behavior and having a favorable personality, generally correspond to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and more recently emphasized in evolutionary psychology models of leadership [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @howell_role_2005]. The importance of prosociality  also corresponds to the @henrich_big_2015 Big-man mechanism, which suggests positive assortment of followers to prosocial leaders can facilitate the evolution of cooperative social norms and prestige-based leadership.
<!-- A few variables did not seem to fit prosocial theme, however, such as *Physical health*, *Physical formidable*, and *Drug consumption* (although drug use is often an important social activity). -->

<!-- DISCUSSES COMPETENCIES FEATURE -->
**Leader capacities**. The individual-level competencies feature was comprised of variables associated with dominance-based leadership, including coercive authority and aggressiveness [@tiger_imperial_1971; @barkow_prestige_1975; @barkow_darwin_1989; @henrich_evolution_2001; @cheng_two_2013] as well as the qualities of knowledgeable or intelligent and experienced or accomplished, traits that have been emphasized by the prestige theory [@henrich_evolution_2001; @cheng_two_2013] and the more general computational services theory [@hagen_leadership_2019; @garfield_evolutionary_2019], and which are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport)). Interestingly, being feared and having supernatural qualities clustered more closely with the prestige-style traits than with the dominance-style traits. There is likely a strong connection between shamanism and ritual-based influence with both prestige and dominance styles of leadership. Ritual and supernatural capacities may represent a dimension of leader qualities in which dominance and prestige are coupled forming an alternative style of leadership, as shamans often provide valued public good, have special expertise and abilities, are feared, and achieve or maintain influence via threats or use of direct and/or indirect aggression. Artistic performance is also implicated in this style of leadership and ceremonial, artistic performances by shamans are common displays which often incorporate the superhuman abilities of shamans [@singh_cultural_2017].

@@

<!-- DISCUSSES SUCCESS FEATURE -->
**Leader success**. Leaders, probably as a consequence of their capacities to provide prosocial benefits (e.g., resolving conflicts and being generous) and/or impose social costs (e.g., punishing and aggressiveness), are also characterized by their material, social, and reproductive success (bottom cluster Figure \@ref(fig:cluster)A), which is consistent with much of the theoretical literature focused on associated outcomes of leadership and social hierarchy [e.g., @von_rueden_mens_2016; @bowles_emergence_2010; @glowacki_role_2013]. Several of these measures of success often co-occur. Among the Trobriands, horticulturalists in Papua New Guinea, @ol06-037 [p. 47] describes relationships between polygynous marriages, increased wealth, and widespread social contacts:

@@

<!-- DISCUSSES BOTH MANAGEMENT FEATURES -->
**Leaders are managers**. There is a history within managerial studies among WEIRD populations of contrasting *leaders* from *managers*. @zaleznik_managers_1977, for example, suggested managers seek systematicity, organization, and stability, whereas leaders are willing to accept a lack of control and creatively navigate uncertainty to produce novel insight and appeale to the emotions of followers. Political anthropology reveals, however, in the absence of state or organizational institutions, local leaders are often tasked with both managing various economic systems while also providing social services and interacting with followers on an inter-personal level [@cohen_comparative_1967]. Within the domain of leader management, there are likely to be two primary features of human leader functions: strategizing and organizing. Our results correspond very closely to what @van_vugt_cognitive_2007 define as *strategic leadership*, where leaders increase the pay-offs of another individuals behavior thereby incentivizing followership versus *coordinating leadership*, where individuals bound by group challenges must effectively organize collective effort.

@@

# Concluding remarks: Beyond the non-WEIRD 
<!-- NOT SURE OF TITLE -- WANT TO EMPHASIZE NON WEIRD DIVERSITY AND ALSO MENTION BACK TO WERID FINDINGS BRIEFLY -->
In recognizing the limitations of most extant psychological, sociological, and managerial data, many social scientists are now diversifying their samples to better understand *human* psychology, behavior, and culture. We caution that the WEIRD people problem in behavioral science cannot be remedied simply by a "non-WEIRD" solution. As our results, as well as much of anthropology and ethnography, demonstrate, within the populations that are now often categorized as non-WEIRD, there is enormous diversity. Furthermore, larger order categories such as "subsistence strategy" are probably less relevant for understanding behavioral diversity than more salient group contexts. Despite limitations, the ethnographic record presents an invaluable opportunity to explore patterns of human diversity and evolutionary theorists and researchers looking beyond WEIRD samples should take special interest in comparative ethnographic methods.

We found leaders in the ethnographic record are portrayed as generally prosocial, serve to resolve conflicts and organize cooperation, and receive material, social, and reproductive benefits. These results conform to the characteristics of leaders that are emphasized in the literature on the evolution of leadership and provide broad cross-cultural support for several influential theories, such as the prestige-dominance model, collective action models, the service-for-prestige model, and the computational services model [for review, see @garfield_evolutionary_2019].

While there is a evidence of a dominace style of leadership involving coercive authority, confidence and aggression, as well as a prestige style involving special knowledge, experience, and innovation, we suggest rethinking dichotomous models of leadership styles and focusing on inter-individual variation in specific leadership competencies. There is likely to be substantial overlap in physical, spiritual, cognitive, and other prosocial endowments and the degree to which a leader relies on any will be shaped by their individual levels of such capacities, the presence and capacities of their rivals, as well as group context, their functions, and cultural norms. For example, widespread preferences for egalitarianism and social equality are likely strong drivers of followers preferences for leaders across diverse societies [@cheng_dominance_2019], but shamanism and ritual based contexts represent a social domain in which ideologies support a style of leadership fusing both dominance and prestige-based competencies. We suspect cultural and pseudoreligous ideologies also play important roles in shaping leadership styles follower preferences among WEIRD populations and in national political movements around the world.
